---
title: "P8105 Homework 2"
author: Heidi Lumish
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(kableExtra)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

## Problem 1

#### Set up the Mr. Trash Wheel data set

Here we will read in the Mr. Trash Wheel Excel sheet and clean the data.

```{r}
trash_wheel_df =
  read_excel(
    "./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N408") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(sports_balls))

```

#### Set up the precipitation data set

Next we will read in and clean the precipitation data for 2018 and 2019.
Then we will combine the 2018 and 2019 precpitation data sets into a new data frame.

```{r}
precipitation_2018 =
  read_excel(
    "./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
    sheet = "2018 Precipitation",
    range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  drop_na(total) %>%
  rename("2018" = total)

precipitation_2019 =
  read_excel(
    "./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
    sheet = "2019 Precipitation",
    range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  drop_na(total) %>%
  rename("2019" = total)

precipitation_df =
  left_join(precipitation_2018, precipitation_2019, by = "month") %>% 
  mutate(month = month.name[month])
```

#### Summarize the Mr. Trash Wheel data set

Mr. Trash Wheel is a "water-wheel vessel that removes trash from the Inner Harbor in Baltimore, Maryland." The Mr. Trash Wheel data set contains information on `r nrow(trash_wheel_df)` dumpsters, with litter collected between `r min(pull(trash_wheel_df, year))` and `r max(pull(trash_wheel_df, year))`. There was a total of `r sum(pull(trash_wheel_df, weight_tons))` tons of litter, with a volume of `r sum(pull(trash_wheel_df, volume_cubic_yards))` cubic yards, removed during this time. The type of litter collected included
`r format(sum(pull(trash_wheel_df, plastic_bottles)), scientific = FALSE)` plastic bottles,
`r format(sum(pull(trash_wheel_df, polystyrene)), scientific = FALSE)` foam containers,
`r sum(pull(trash_wheel_df, glass_bottles))` glass bottles,
`r format(sum(pull(trash_wheel_df, grocery_bags)), scientific = FALSE)` grocery bags,
`r format(sum(pull(trash_wheel_df, chip_bags)), scientific = FALSE)` chip bags, and
`r sum(pull(trash_wheel_df, sports_balls))` sports balls.
The median number of sports balls collected each year was
`r median(pull(trash_wheel_df, sports_balls)[pull(trash_wheel_df, year) == 2014])` in 2014,
`r median(pull(trash_wheel_df, sports_balls)[pull(trash_wheel_df, year) == 2015])` in 2015,
`r median(pull(trash_wheel_df, sports_balls)[pull(trash_wheel_df, year) == 2016])` in 2016,
`r median(pull(trash_wheel_df, sports_balls)[pull(trash_wheel_df, year) == 2017])` in 2017,
`r median(pull(trash_wheel_df, sports_balls)[pull(trash_wheel_df, year) == 2018])` in 2018, and
`r median(pull(trash_wheel_df, sports_balls)[pull(trash_wheel_df, year) == 2019])` in 2019.

#### Summarize the precipitation data set

The precipitation dataset summarizes the precipitation during `r nrow(precipitation_df)` months in 2018 and 2019. The total precipitation between January and December 2018 was `r sum(pull(precipitation_df, "2018"))` inches. The median monthly precipitation was `r median(pull(precipitation_df, "2018"))` inches. The maximum monthly precipitation in 2018 was `r max(pull(precipitation_df, "2018"))` inches and occurred during the month of September. For 2019, precipitation data is available only between January and June. The total precipitation in 2019 was `r sum(pull(precipitation_df, "2019"), na.rm = TRUE)` inches. The median monthly precipitation was `r median(pull(precipitation_df, "2019"), na.rm = TRUE)` and the maximum monthly precipitation was `r max(pull(precipitation_df, "2019"), na.rm = TRUE)` inches and occurred in March.

## Problem 2

#### Import and clean the pols-month file

First, we will import and clean the data in pols-month.csv. Below is the code to convert the variable "mon" to separate "day," "month," and "year" variables. In addition, the code below creates a new variable "president" with levels "gop" and "dem," and then removes the prez_dem, prez_gop, and day variables.

```{r, message = FALSE}
pols_df = read_csv("./data/pols-month.csv") %>% 
  janitor::clean_names() %>% 
  separate(mon,
           c("year", "month", "day"),
           sep = "([-])",
           convert = TRUE) %>% 
  mutate(month = month.name[month]) %>% 
  mutate(president = ifelse(prez_dem == "1", "dem", "gop")) %>% 
  select(-prez_dem, -prez_gop, -day) %>% 
  relocate(year, month, president)
```

#### Import and clean the snp file

Next we will import and clean the data in the snp.csv file, following similar methods to the pols-month file.

```{r, message = FALSE}
snp_df = read_csv("./data/snp.csv") %>% 
  janitor::clean_names() %>%
  mutate(date = mdy(date)) %>% 
  separate(date,
           c("year", "month", "day"),
           sep = "([-])",
           convert = TRUE) %>%
  mutate(year = ifelse(
    year>2021, year - 100, year)) %>% 
  mutate(month = month.name[month]) %>% 
  select(-day) %>%
  relocate(year, month)
```

#### Import and clean the unmployment file

Next we will import and clean the data in the unemployment file. 

```{r, message = FALSE}
unemployment_df = read_csv("./data/unemployment.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "percent"
  )%>% 
  mutate(
    month =
           case_when(
             month == "jan" ~ "January",
             month == "feb" ~ "February",
             month == "mar" ~ "March",
             month == "apr" ~ "April",
             month == "may" ~ "May",
             month == "jun" ~ "June",
             month == "jul" ~ "July",
             month == "aug" ~ "August",
             month == "sep" ~ "September",
             month == "oct" ~ "October",
             month == "nov" ~ "November",
             month == "dec" ~ "December"))
```

#### Merge the three data sets: pols, snp, and unemployment

Finally, we will merge the three data sets together.

```{r}
five_thirty_eight_df =
  left_join(pols_df, snp_df, by = c("year", "month")) %>% 
              left_join(., unemployment_df, by = c("year", "month"))
```

#### Summary of the FiveThirtyEight data set

The FiveThirtyEight data set combines three files: pols-month, snp, and unemployment. The pols-month dataset included `r nrow(pols_df)` observations of `r ncol(pols_df)` variables, with data collected between `r min(pull(pols_df, "year"))` and `r max(pull(pols_df, "year"))`. It incorporated information on the number of democratic and republican presidents, governors, senators, and representatives on a given date, based on polling information.

The snp dataset included `r nrow(snp_df)` observations of `r ncol(snp_df)` variables, with data collected between `r min(pull(snp_df, "year"))` and `r max(pull(snp_df, "year"))`. The data provided information about the Standard & Poor's stock market index, specifically the closing values of the S&P stock index on the associated date. The lowest S&P closing value during this period was `r round(min(pull(snp_df, "close"), na.rm = TRUE), digits = 2)`, the highest closing value was `r round(max(pull(snp_df, "close"), na.rm = TRUE), digits = 2)` and the median closing value was `r round(median(pull(snp_df, "close"), na.rm = TRUE), digits = 2)`.

The unemployment data set included `r nrow(unemployment_df)` observations of `r ncol(unemployment_df)` variables, with data collected between `r min(pull(unemployment_df, "year"))` and `r max(pull(unemployment_df, "year"))`. The data provide information on the percentage of unemployment per month. The lowest unemployment rate during this period was `r min(pull(five_thirty_eight_df, "percent"), na.rm = TRUE)`%, the highest rate was `r max(pull(five_thirty_eight_df, "percent"), na.rm = TRUE)`% and the median unemployment rate was `r median(pull(five_thirty_eight_df, "percent"), na.rm = TRUE)`%.

The combined FiveThirtyEight data set has `r nrow(five_thirty_eight_df)` observations and `r ncol(five_thirty_eight_df)` columns, and it includes data collected between `r min(pull(five_thirty_eight_df, "year"))` and `r max(pull(five_thirty_eight_df, "year"))`. 

## Problem 3

First we will load and clean the Baby Names data set.

```{r, message = FALSE}
baby_names_df = read_csv("./data/Popular_Baby_Names.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    childs_first_name = str_to_title(childs_first_name),
    ethnicity =
           case_when(
             ethnicity == "ASIAN AND PACIFIC ISLANDER" ~ "Asian and Pacific Islander",
             ethnicity == "ASIAN AND PACI" ~ "Asian and Pacific Islander",
             ethnicity == "WHITE NON HISPANIC" ~ "Non-Hispanic White",
             ethnicity == "WHITE NON HISP" ~ "Non-Hispanic White",
             ethnicity == "HISPANIC" ~ "Hispanic",
             ethnicity == "BLACK NON HISP" ~ "Non-Hispanic Black",
             ethnicity == "BLACK NON HISPANIC" ~ "Non-Hispanic Black")
  ) %>%
  distinct()
```

#### Table of most popular baby names for girls

This code chunk will produce a table showing the rank in popularity of the name “Olivia” as a female baby name over time, by ethnicity.

```{r}
olivia_df = baby_names_df %>%
  filter(childs_first_name == "Olivia") %>% 
  select(-gender, -childs_first_name, -count) %>% 
  arrange(year_of_birth) %>% 
  pivot_wider(
    names_from = "year_of_birth",
    values_from = "rank"
  ) %>% 
  arrange(ethnicity)

colnames(olivia_df) = str_to_title(colnames(olivia_df))

knitr::kable(
  olivia_df[1:4, ],
  caption = ,
  align = "lcccccc")
```

#### Table of most popular baby names for boys

This code chunk will produce a table showing the most popular boys name (ranked #1) by ethnicity over time:

```{r}
boy_names = baby_names_df %>%
  filter(gender == "MALE", rank == 1) %>% 
  select(-gender, -rank, -count) %>% 
  arrange(year_of_birth) %>% 
  pivot_wider(
    names_from = "year_of_birth",
    values_from = "childs_first_name"
  ) %>% 
  arrange(ethnicity)

colnames(boy_names) = str_to_title(colnames(boy_names))

knitr::kable(
  boy_names[1:4, ],
  caption = ,
  align = "lcccccc")
```

#### Scatterplot

The following code chunk can be used to generate a scatterplot showing the number of children with a name against the rank in popularity of that name, among male, white, non-Hispanic children born in 2016.

```{r}
white_boy_names = baby_names_df %>%
  filter(gender == "MALE", ethnicity == "Non-Hispanic White", year_of_birth == "2016") 

ggplot(white_boy_names, aes(x = rank, y = count)) +
  geom_point(color = "blue") +
  labs(
    title = "Frequency of baby name vs. popularity of name",
    x = "Rank of baby name",
    y = "Frequency of baby name")+
  theme_minimal()
```
